---
title: "DeltaMultipleBenefits"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DeltaMultipleBenefits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The functions contained in the DeltaMultipleBenefits package are intended to facilitate estimation of the net impacts of scenarios of landscape change in the Sacramento-San Joaquin River Delta. It provides tools for applying included data and species distribution models to user-supplied landscapes to estimate a range of benefits to the Delta community: Agricultural Livelihoods, Water Quality, Climate Change Resilience, and Biodiversity Support. By comparing metrics from proposed scenarios of landscape change to metrics estimated for a baseline landscape representing current conditions, the estimated net change in each metric helps identify synergies and trade-offs among multiple benefits. The results provide insights into the direction and expected magnitude of change in each metric, are intended to help establish a common understanding among members of the Delta community and facilitate land use planning, management, conservation, and community engagement.

This tutorial outlines the major steps of analyzing a new scenario of landscape change or reproducing the results in Dybala et al. (2023), including:
1. Preparing a new scenario
2. Summarizing the total extent of each land cover
3. Estimating the total landscape scores for each metric
4. Estimating the change from baseline scores

```{r setup}
library(DeltaMultipleBenefits)
```

# 1. Preparing a new scenario

From polygons --> map to key --> rasterize
Or use our baseline/scenario landscapes (CDFW BIOS)

# 2. Summarizing the total extent of each land cover


# 3. Estimating the total landscape scores for each metric

data required:
- species models and spp key (in R package)
- landcover-specific metrics
- baseline scores (and other scenario scores for comparison)

for biodiversity support: generate predictors for spatial models --> fit spatial models and produce predicted distributions (with/without thresholds) -- compare to our results (CDFW BIOS)

# 4. Estimating the change from baseline scores

repeat process above for your own baseline, or use our baseline/scenario scores (provide in R package)
